---
title: "Predictive Checks"
author: "Bryan Roemelt"
date: "2026-05-12"
output: html_document
---

# ── Setup ─────────────────────────────────────────────────────────────────────
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, here, posterior, cmdstanr, patchwork, mvtnorm)

draws_list <- readRDS(here("empirical", "empirical_draws.rds"))
sim_draws_list <- readRDS(here("simulation", "simulated_draws.rds"))
```

## ── Kalman filter in R ───────────────────────────────────────────────────────
```{r}
run_kalman <- function(r_value, q_value, stan_data) {
  ntrials     <- stan_data$ntrials
  nfeatures   <- stan_data$nfeatures
  ncategories <- stan_data$ncategories
  mu          <- stan_data$initial_mu
  sigma       <- lapply(1:ncategories, function(c)
    diag(stan_data$initial_sigma_diag, nfeatures))
  r_mat       <- diag(r_value, nfeatures)
  q_mat       <- diag(q_value, nfeatures)
  I_mat       <- diag(nfeatures)
  preds       <- integer(ntrials)

  for (i in seq_len(ntrials)) {
    # Predict
    sigma <- lapply(sigma, function(s) s + q_mat)

    # Decide
    x         <- stan_data$obs[i, ]
    log_probs <- sapply(1:ncategories, function(c)
      dmvnorm(x, mean = mu[[c]], sigma = sigma[[c]] + r_mat, log = TRUE))
    probs     <- exp(log_probs) / sum(exp(log_probs))
    preds[i]  <- sample(1:ncategories, 1, prob = probs)

    # Update true category
    c_true        <- stan_data$cat_true[i]
    innov         <- x - mu[[c_true]]
    S             <- sigma[[c_true]] + r_mat
    K             <- sigma[[c_true]] %*% solve(S)
    IK            <- I_mat - K
    mu[[c_true]]  <- mu[[c_true]] + K %*% innov
    sigma[[c_true]] <- IK %*% sigma[[c_true]] %*% t(IK) + K %*% r_mat %*% t(K)
    sigma[[c_true]] <- 0.5 * (sigma[[c_true]] + t(sigma[[c_true]]))
  }
  preds
}
```


# ── Empirical checks ──────────────────────────────────────────────────────────

## ── Stan data builder for empirical ──────────────────────────────────────────
```{r}
make_stan_data <- function(id) {
  participant_data <- transformed_data %>% filter(subject == id)
  list(
    ntrials            = nrow(participant_data),
    nfeatures          = 5,
    ncategories        = 4,
    cat_true           = participant_data$category,
    y                  = participant_data$response,
    obs                = as.matrix(participant_data[, c("eyes", "buffness", "arms", "spots", "color")]),
    initial_mu         = lapply(1:4, function(c) rep(0.5, 5)),
    initial_sigma_diag = 1.0,
    prior_logr_mean    = log(0.2),
    prior_logr_sd      = 0.5,
    prior_logq_mean    = log(0.01),
    prior_logq_sd      = 0.5
  )
}
```

## ── Prior predictive ─────────────────────────────────────────────────────────
```{r}
generate_prior_ppc <- function(stan_data, participant_id, n_draws = 500) {
  logr <- rnorm(n_draws, log(0.2), 0.5)
  logq <- rnorm(n_draws, log(0.01), 0.5)

  curves <- map_dfr(seq_len(n_draws), function(s) {
    preds   <- run_kalman(exp(logr[s]), exp(logq[s]), stan_data)
    correct <- as.integer(preds == stan_data$cat_true)
    tibble(
      sample_id = s,
      trial     = seq_len(stan_data$ntrials),
      correct   = correct
    ) |> mutate(cum_acc = cumsum(correct) / row_number())
  })

  curves |>
    group_by(trial) |>
    summarise(
      q05 = quantile(cum_acc, 0.05),
      q25 = quantile(cum_acc, 0.25),
      q50 = median(cum_acc),
      q75 = quantile(cum_acc, 0.75),
      q95 = quantile(cum_acc, 0.95),
      .groups = "drop"
    ) |>
    mutate(
      obs = cumsum(stan_data$y == stan_data$cat_true) / seq_len(stan_data$ntrials),
      ID  = participant_id
    )
}
```

## ── Posterior predictive ─────────────────────────────────────────────────────
```{r}
generate_post_ppc <- function(draws, stan_data, participant_id) {
  # Extract p[i] draws — one row per draw, one column per trial
  p_cols  <- grep("^p\\[", names(draws), value = TRUE)
  p_draws <- as.matrix(draws[, p_cols])  # n_draws x (ntrials * ncategories)

  # Extract per-trial probability vectors and simulate responses
  ntrials     <- stan_data$ntrials
  ncategories <- stan_data$ncategories

  yrep <- t(sapply(seq_len(nrow(p_draws)), function(d) {
    sapply(seq_len(ntrials), function(i) {
      cols  <- seq((i - 1) * ncategories + 1, i * ncategories)
      probs <- as.numeric(p_draws[d, cols])
      sample(1:ncategories, 1, prob = probs)
    })
  }))

  # Cumulative accuracy per draw
  yrep_correct  <- sweep(yrep, 2, stan_data$cat_true, "==") + 0
  yrep_cum_acc  <- t(apply(yrep_correct, 1,
                           function(r) cumsum(r) / seq_along(r)))

  obs_cum_acc <- cumsum(stan_data$y == stan_data$cat_true) / seq_len(ntrials)

  tibble(
    trial = seq_len(ntrials),
    q05   = apply(yrep_cum_acc, 2, quantile, 0.05),
    q25   = apply(yrep_cum_acc, 2, quantile, 0.25),
    q50   = apply(yrep_cum_acc, 2, quantile, 0.50),
    q75   = apply(yrep_cum_acc, 2, quantile, 0.75),
    q95   = apply(yrep_cum_acc, 2, quantile, 0.95),
    obs   = obs_cum_acc,
    ID    = participant_id
  )
}
```

## ── Generate predictions with progress tracker ───────────────────────────────
```
prior_ppc <- imap_dfr(draws_list, function(draws, id)
  generate_prior_ppc(make_stan_data(id), participant_id = id))

post_ppc <- imap_dfr(draws_list, function(draws, id)
  generate_post_ppc(draws, make_stan_data(id), participant_id = id))
```

```{r}
n_subjects <- length(draws_list)

cat("Generating prior predictive checks...\n")
prior_ppc <- imap_dfr(draws_list, function(draws, id) {
  cat(sprintf("  Prior  - participant %s / %d\n", id, n_subjects))
  generate_prior_ppc(make_stan_data(id), participant_id = id)
})

cat("Generating posterior predictive checks...\n")
post_ppc <- imap_dfr(draws_list, function(draws, id) {
  cat(sprintf("  Posterior - participant %s / %d\n", id, n_subjects))
  generate_post_ppc(draws, make_stan_data(id), participant_id = id)
})

cat("Done.\n")
```

## ── Plot empirical checks ────────────────────────────────────────────────────
```{r}
p_emp_prior <- ggplot(prior_ppc, aes(x = trial)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#c97b6e", alpha = 0.20) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#c97b6e", alpha = 0.40) +
  geom_line(aes(y = q50), color = "#a85a4a", linewidth = 1) +
  geom_line(aes(y = obs, group = ID), color = "#333333",
            linewidth = 0.5, alpha = 0.3) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "grey50") +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(
    title    = "Prior Predictive Check: Empirical Data",
    subtitle = "Ribbons: 50% and 90% prior predictive intervals | Grey: simulated agents",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  theme_classic()

p_emp_post <- ggplot(post_ppc, aes(x = trial)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#6e8fc9", alpha = 0.20) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#6e8fc9", alpha = 0.40) +
  geom_line(aes(y = q50), color = "#4a6fa5", linewidth = 1) +
  geom_line(aes(y = obs, group = ID), color = "#333333",
            linewidth = 0.5, alpha = 0.3) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "grey50") +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(
    title    = "Posterior Predictive Check: Empirical Data",
    subtitle = "Ribbons: 50% and 90% posterior predictive intervals | Grey: participant observations",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  theme_classic()

p_emp_prior
p_emp_post
```

## ── Numerical summaries ──────────────────────────────────────────────────────
```{r}
bind_rows(
  prior_ppc %>% mutate(check = "Prior"),
  post_ppc  %>% mutate(check = "Posterior")
) %>%
  group_by(check) %>%
  summarise(
    mean_width = mean(q95 - q05),
    coverage   = mean(obs >= q05 & obs <= q95),
    mae        = mean(abs(q50 - obs))
  )
```

# ── Simulation checks ─────────────────────────────────────────────────────────

## ── Stan data builder for simulation ─────────────────────────────────────────

```{r}
make_stan_data_sim <- function(id) {
  participant_data <- sim_data %>% filter(subject_id == id)
  list(
    ntrials            = nrow(participant_data), #96
    nfeatures          = 5,
    ncategories        = 4,
    cat_true           = participant_data$feedback,
    y                  = participant_data$choice,
    obs                = as.matrix(participant_data[, c("eyes", "buffness", "arms", "spots", "color")]),
    initial_mu         = lapply(1:4, function(c) rep(0.5, 5)),
    initial_sigma_diag = 1.0,
    prior_logr_mean    = 0,
    prior_logr_sd      = 1,
    prior_logq_mean    = -2,
    prior_logq_sd      = 1
  )
}
```

## ── Prior predictive ─────────────────────────────────────────────────────────

```{r}
generate_prior_ppc <- function(stan_data, participant_id, n_draws = 500) {
  logr <- rnorm(n_draws, 0, 1)
  logq <- rnorm(n_draws, -2, 1)

  curves <- map_dfr(seq_len(n_draws), function(s) {
    preds   <- run_kalman(exp(logr[s]), exp(logq[s]), stan_data)
    correct <- as.integer(preds == stan_data$cat_true)
    tibble(
      sample_id = s,
      trial     = seq_len(stan_data$ntrials),
      correct   = correct
    ) |> mutate(cum_acc = cumsum(correct) / row_number())
  })

  curves |>
    group_by(trial) |>
    summarise(
      q05 = quantile(cum_acc, 0.05),
      q25 = quantile(cum_acc, 0.25),
      q50 = median(cum_acc),
      q75 = quantile(cum_acc, 0.75),
      q95 = quantile(cum_acc, 0.95),
      .groups = "drop"
    ) |>
    mutate(
      obs = cumsum(stan_data$y == stan_data$cat_true) / seq_len(stan_data$ntrials),
      ID  = participant_id
    )
}
```

## ── Posterior predictive ─────────────────────────────────────────────────────

```{r}
generate_post_ppc <- function(draws, stan_data, participant_id) {
  # Extract p[i] draws — one row per draw, one column per trial
  p_cols  <- grep("^p\\[", names(draws), value = TRUE)
  p_draws <- as.matrix(draws[, p_cols])  # n_draws x (ntrials * ncategories)

  # Extract per-trial probability vectors and simulate responses
  ntrials     <- stan_data$ntrials
  ncategories <- stan_data$ncategories

  yrep <- t(sapply(seq_len(nrow(p_draws)), function(d) {
    sapply(seq_len(ntrials), function(i) {
      cols  <- seq((i - 1) * ncategories + 1, i * ncategories)
      probs <- as.numeric(p_draws[d, cols])
      sample(1:ncategories, 1, prob = probs)
    })
  }))

  # Cumulative accuracy per draw
  yrep_correct  <- sweep(yrep, 2, stan_data$cat_true, "==") + 0
  yrep_cum_acc  <- t(apply(yrep_correct, 1,
                           function(r) cumsum(r) / seq_along(r)))

  obs_cum_acc <- cumsum(stan_data$y == stan_data$cat_true) / seq_len(ntrials)

  tibble(
    trial = seq_len(ntrials),
    q05   = apply(yrep_cum_acc, 2, quantile, 0.05),
    q25   = apply(yrep_cum_acc, 2, quantile, 0.25),
    q50   = apply(yrep_cum_acc, 2, quantile, 0.50),
    q75   = apply(yrep_cum_acc, 2, quantile, 0.75),
    q95   = apply(yrep_cum_acc, 2, quantile, 0.95),
    obs   = obs_cum_acc,
    ID    = participant_id
  )
}
```

## ── Generate predictions (run once) ──────────────────────────────────────────

```
sim_prior_ppc <- imap_dfr(sim_draws_list, function(draws, id)
  generate_prior_ppc(make_stan_data_sim(id), participant_id = id))

sim_post_ppc <- imap_dfr(sim_draws_list, function(draws, id)
  generate_post_ppc(draws, make_stan_data_sim(id), participant_id = id))
```

```{r}
n_subjects <- length(sim_draws_list)

cat("Generating prior predictive checks...\n")
sim_prior_ppc <- imap_dfr(sim_draws_list, function(draws, id) {
  cat(sprintf("  Prior  - participant %s / %d\n", id, n_subjects))
  generate_prior_ppc(make_stan_data_sim(id), participant_id = id)
})

cat("Generating posterior predictive checks...\n")
sim_post_ppc <- imap_dfr(sim_draws_list, function(draws, id) {
  cat(sprintf("  Posterior - participant %s / %d\n", id, n_subjects))
  generate_post_ppc(draws, make_stan_data_sim(id), participant_id = id)
})

cat("Done.\n")
```

## ── Plot simulation checks ───────────────────────────────────────────────────

```{r}
p_sim_prior <- ggplot(sim_prior_ppc, aes(x = trial)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#c97b6e", alpha = 0.20) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#c97b6e", alpha = 0.40) +
  geom_line(aes(y = q50), color = "#a85a4a", linewidth = 1) +
  geom_line(aes(y = obs, group = ID), color = "#333333",
            linewidth = 0.5, alpha = 0.3) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "grey50") +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(
    title    = "Prior Predictive Check: Simulated Data",
    subtitle = "Ribbons: 50% and 90% prior predictive intervals | Grey: simulated agents",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  theme_classic()

p_sim_post <- ggplot(sim_post_ppc, aes(x = trial)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#6e8fc9", alpha = 0.20) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#6e8fc9", alpha = 0.40) +
  geom_line(aes(y = q50), color = "#4a6fa5", linewidth = 1) +
  geom_line(aes(y = obs, group = ID), color = "#333333",
            linewidth = 0.5, alpha = 0.3) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "grey50") +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(
    title    = "Posterior Predictive Check: Simulated Data",
    subtitle = "Ribbons: 50% and 90% posterior predictive intervals | Grey: simulated agents",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  theme_classic()

p_sim_prior
p_sim_post
```

## ── Numerical summaries ──────────────────────────────────────────────────────
```{r}
bind_rows(
  sim_prior_ppc %>% mutate(check = "Prior"),
  sim_post_ppc  %>% mutate(check = "Posterior")
) %>%
  group_by(check) %>%
  summarise(
    mean_width = mean(q95 - q05),
    coverage   = mean(obs >= q05 & obs <= q95),
    mae        = mean(abs(q50 - obs))
  )
```
