---
title: "data_exploration"
author: "Christian Stenbro"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
pacman::p_load(tidyverse, here, tidyverse, posterior, cmdstanr, brms, bayesplot, stringr, dplyr, ggplot2)
```

```{r}
data <- read_delim(here("empirical/AlienData.txt"))
```


```{r}
transformed_data <- data %>% select(condition, 
                                    subject, 
                                    trial,
                                    session,
                                    cycle,
                                    category, 
                                    response,
                                    stimulus) %>%
                              filter(cycle != 0) %>% 
                              filter(condition == 1) %>% 
                              mutate(subject = ifelse(condition == 1, subject, subject + 26)) %>% 
                              mutate(stimulus = str_remove(stimulus, "\\.jpg$"),
                                     eyes = str_sub(stimulus, 1, 1),
                                     buffness = str_sub(stimulus, 2, 2),
                                     arms = str_sub(stimulus, 3, 3),
                                     spots = str_sub(stimulus, 4, 4),
                                     color = str_sub(stimulus, 5, 5)
                                    ) %>% 
                              mutate(across(c(eyes, buffness, arms, spots, color), as.numeric)) %>% 
                              select(-stimulus)
```


# understanding the variables
```{r}
summary(transformed_data)
```

What are these variables? 

- competence
- motivation
- communication
- complement

For now, we will not use them in the model.

There are also these variables:

- condition
- session 
- cycle

Which might be (probably are) relevant for the modelling. I wonder how we can learn more about them?

Notice: "You should then visually explore the empirical data and fit the Stan model to them. **Fitting the model by participant and sticking to 1 session** is sufficient to pass."


```{r}
length(unique(transformed_data$subject))
```

# data visualization

How can we start exploring this data?

Perhaps a good first step would be to visualize, for a few round, how a participant updates their responses based on the feedback.

Potential problem: We are modelling a four-option response (tracking two binary dimensions) as a two-option response. How do we go about this practically? Did the participants receive direct feedback on whether their categorization was correct on both dimensions, or just a black-box correct/incorrect feedback?
```{r}
transformed_data %>%
  group_by(subject, session) %>%
  summarise(n = n()) %>%
  arrange(desc(n))
```


```{r}
accuracy_data <- transformed_data %>%
  mutate(correct = as.numeric(category == response)) %>%
  group_by(subject, session) %>%
  summarise(
    accuracy = mean(correct, na.rm = TRUE),
    .groups = "drop"
  )

group_means <- accuracy_data %>%
  group_by(session) %>%
  summarise(mean_accuracy = mean(accuracy, na.rm = TRUE))
```

```{r}
ggplot(filter(accuracy_data, session == 1),
       aes(x = factor(subject), y = accuracy)) +
  geom_col(alpha = 0.5) +

  # chance level
  geom_hline(yintercept = 0.25, linetype = "dashed") +

  # group mean
  geom_hline(
    yintercept = group_means$mean_accuracy[group_means$session == 1],
    color = "red",
    linewidth = 1
  ) +

  labs(
    x = "Subject",
    y = "Accuracy (%)"
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  theme_classic()

```
```{r}
ggplot(filter(accuracy_data, session == 2),
       aes(x = factor(subject), y = accuracy)) +
  geom_col(alpha = 0.5) +
  geom_hline(yintercept = 0.25, linetype = "dashed") +
  geom_hline(
    yintercept = group_means$mean_accuracy[group_means$session == 2],
    color = "red",
    linewidth = 1
  ) +
  labs(
    x = "Subject",
    y = "Accuracy (%)"
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  theme_classic()
```

```{r}
ggplot(filter(accuracy_data, session == 3),
       aes(x = factor(subject), y = accuracy)) +
  geom_col(alpha = 0.5) +
  geom_hline(yintercept = 0.25, linetype = "dashed") +
  geom_hline(
    yintercept = group_means$mean_accuracy[group_means$session == 3],
    color = "red",
    linewidth = 1
  ) +
  labs(
    x = "Subject",
    y = "Accuracy (%)"
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  theme_classic()
```

