---
title: "data_exploration"
author: "Christian Stenbro"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
pacman::p_load("here", "tidyverse")
```

```{r}
# loading data
data <- read_delim("/Users/christianstenbro/AU/MSc/ACM/assignment_4_LOCAL/AlienData.txt")
```

```{r}
data
```

We can remove the .jpg extension from the stimulus column (which is the feature binary codes).

```{r}
data$stimulus <- str_replace(data$stimulus, ".jpg", "")
```

# understanding the variables
```{r}
summary(data)
```

What are these variables? 

- competence
- motivation
- communication
- complement

For now, we will not use them in the model.

There are also these variables:

- condition
- session 
- cycle

Which might be (probably are) relevant for the modelling. I wonder how we can learn more about them?

Notice: "You should then visually explore the empirical data and fit the Stan model to them. **Fitting the model by participant and sticking to 1 session** is sufficient to pass."

# data subset

First, we select only the relevant columns and remove data that are not session 1: 

```{r}
data_subset <- data %>% select(condition, 
                               subject, 
                               session,
                               cycle,
                               trial,
                               test,
                               stimulus,
                               category,
                               response,
                               dangerous,
                               nutricious,
                               correct) %>% filter(session == 1)
```

```{r}
data_subset
```

```{r}
length(unique(data_subset$subject))
```

# data visualization

How can we start exploring this data?

Perhaps a good first step would be to visualize, for a few round, how a participant updates their responses based on the feedback.

Potential problem: We are modelling a four-option response (tracking two binary dimensions) as a two-option response. How do we go about this practically? Did the participants receive direct feedback on whether their categorization was correct on both dimensions, or just a black-box correct/incorrect feedback?

